Implications of taxonomic bias for microbial differential-abundance analysis

### 2022-08-19 differential-abundance.Rmd
Make edits from AW and BC

Taxonomic bias affects the measured FDs in a species' absolute abundance in a similar fashion.
First, consider a species' abundance measurement $\tilde A_i^{(a)}$ derived from a non-MGS measurement of total-community abundance using Equation \@ref(eq:density-prop-meas).
Equation \@ref(eq:density-prop-error) indicates that the FE in this measurement equals the (constant) species efficiency multiplied by ${\bar B^{\mtot (a)}}/{\bar B^{(a)}}$, the ratio of the mean efficiency of the total-abundance measurement to that of the MGS measurement.
This ratio can vary, creating error in the measured FD between samples.
Notably, if the FDs in the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement, then the two offset each other and lead to more stable FEs (and hence more accurate FDs) in $\tilde A_i^{(a)}$.
We discuss how this possibility might be exploited in real experimental workflows in Section \@ref(solutions).

Now consider a species' abundance derived from a reference species using Equation \@ref(eq:density-ratio-meas).
Equation \@ref(eq:density-ratio-error) indicates that the FE equals a constant ratio in species' efficiencies multiplied by the FE in the measurement of the reference species' abundance.
If the abundance of the reference species can determined up to constant FE, then the FE in $\tilde A_i^{(a)}$ will also be constant, leading to accurate FDs.

@@

Although researchers typically use more sophisticated models, the simple linear regression model in Equation \@ref(eq:regression) provides an intuitive basis for understanding the effect of taxonomic bias on regression-based DA analyses.
Appendix \@ref(appendix-regression) describes a general framework for finding the error in the estimated slope coefficient $\beta$ caused by taxonomic bias; here, we summarize and illustrate these results for the various abundance measures decribed in Section \@ref(abundance-measurement).
In each case, a multiplicative DA analysis can be conducted by setting $y$ equal to the logarithm of the untransformed abundance measurement ($\tilde P_i$, $\tilde R_{i/j}$, or $\tilde A_i$).
Consider an abundance measure such as the ratio between species $i$ and $j$, $\tilde R_{i/j}$.
Taxonomic bias creates a constant FE in $\tilde R_{i/j}$ that translates to a constant additive shift in $\log \tilde R_{i/j}$ and therefore does not affect the estimate of the slope $\hat \beta$.

Now consider the absolute abundance of a species $i$ obtained by normalizing its proportion to an non-MGS measurement of total community abundance, which for illustration we assume to be completely accurate.
The measured abundance $\tilde A_i$ thus has a FE that varies inversely with the mean efficiency $\bar B$.
If $\bar B$ is stable across samples, then the additive error in $\log \tilde A_i$ will be approximately constant and $\hat \beta$ is again unaffected---bias has no effect on the DA result.
If, on the other hand, $\log \bar B$ tends to linearly vary with the covariate $x$, then the additive error in $\log \tilde A_i$ will linearly vary in the opposite direction and cause a systematic error in $\hat \beta$ that is equal in magnitude but opposite in sign to the slope of $\log \bar B$ with $x$.
A third possibility is that $\bar B$ varies in an essentially random manner that is independent of the covariate $x$.
In this scenario, the error from taxonomic bias in the measurement $\log \tilde A_{i}$ acts to increase the noise in $\hat beta$ (i.e., increase its standard error), but does not cause a systematic error.

>>>>

Taxonomic bias affects measured FDs in a species' absolute abundance in a similar fashion.
First, consider a species' abundance measurement $\tilde A_i^{(a)}$ derived from a non-MGS measurement of total-community abundance using Equation \@ref(eq:density-prop-meas).
Equation \@ref(eq:density-prop-error) indicates that the FE in this measurement equals the (constant) species efficiency multiplied by ${\bar B^{\mtot (a)}}/{\bar B^{(a)}}$, the ratio of the mean efficiency of the total-abundance measurement to that of the MGS measurement.
This ratio can vary, creating error in the measured FD between samples.
A notable special case is if the FDs in the mean efficiency of the total-abundance measurement mirrors that of the MGS measurement; then, the two will offset each other and lead to more stable FEs (and hence more accurate FDs) in $\tilde A_i^{(a)}$.
We discuss how this possibility might be exploited in real experimental workflows in Section \@ref(solutions).

Now consider a species' abundance derived from a reference species using Equation \@ref(eq:density-ratio-meas).
Equation \@ref(eq:density-ratio-error) indicates that the FE equals a constant ratio in species' efficiencies multiplied by the FE in the measurement of the reference species' abundance.
If the abundance of the reference species can be determined up to constant FE across samples, then the FE in $\tilde A_i^{(a)}$ will also be constant, leading to accurate FDs.

@@

Although researchers often use more sophisticated models, the simple linear regression model in Equation \@ref(eq:regression) provides an intuitive basis for understanding the effect of taxonomic bias on regression-based DA analyses.
Appendix \@ref(appendix-regression) describes a general framework for finding the error in the estimated slope coefficient $\beta$ caused by taxonomic bias; here, we summarize and illustrate these results for the various abundance measures described in Section \@ref(abundance-measurement).
In each case, a multiplicative DA analysis can be conducted by setting $y$ equal to the logarithm of the untransformed abundance measurement ($\tilde P_i$, $\tilde R_{i/j}$, or $\tilde A_i$).
Consider an abundance measure such as the ratio between species $i$ and $j$, $\tilde R_{i/j}$.
Taxonomic bias creates a constant FE in $\tilde R_{i/j}$ that translates to a constant additive shift in $\log \tilde R_{i/j}$ and therefore does not affect the estimate of the slope $\hat \beta$.

Now consider the absolute abundance of a species $i$ obtained by normalizing its proportion to an non-MGS measurement of total community abundance, which for illustration we assume to be completely accurate.
The measured abundance $\tilde A_i$ thus has a FE that varies inversely with the mean efficiency $\bar B$.
If $\bar B$ is stable across samples, then the additive error in $\log \tilde A_i$ will be approximately constant and $\hat \beta$ is again unaffected---bias has no effect on the DA result.
If, on the other hand, $\log \bar B$ tends to linearly vary with the covariate $x$, then the additive error in $\log \tilde A_i$ will linearly vary in the opposite direction and cause a systematic error in $\hat \beta$ that is equal in magnitude but opposite in sign to the slope of $\log \bar B$ with $x$.
A third possibility is that $\bar B$ varies in an essentially random manner that is independent of the covariate $x$.
In this scenario, the error from taxonomic bias in the measurement $\log \tilde A_{i}$ acts to increase the noise in $\hat \beta$ (i.e., increase its standard error), but does not cause a systematic error.

### 2022-08-19 introduction.Rmd
Make edits from AW and BC

<!-- P: Taxonomic bias and DA -->
The standard approach to countering taxonomic bias is to standardize the measurement protocol used within a given study.
Statistical analyses are then conducted with the tacit assumption that all samples will be affected by bias in the same way and so the differences between samples will be unaffected.
This argument is at least intuitively plausible for DA analyses based on multiplicative or fold differences (FDs) in a taxon's abundance.
If bias caused a species' abundance to be consistently measured as 10-fold greater than its actual value, then we would still recover the correct FDs among samples.
<!-- (@kevorkian2018esti, @lloyd2020evid). -->
However, @mclaren2019cons showed mathematically and with MGS measurements of artificially constructed ('mock') communities that consistent taxonomic bias can create fold errors (FEs) that vary across samples and, as a result, majorly distort cross-sample comparisons.
In particular, they showed that the FE in a species' proportion---the most common measure of relative abundance---varies among samples, distorting the observed FDs between samples.
In some cases, bias can even lead to incorrect inferences about the direction of change (for example, by causing a taxon that decreased to appear to increase).
Yet @mclaren2019cons also found that other abundance measures---those based on the ratios among species---have proportional errors and may lead to more robust DA analyses.
The implications of these findings for DA analysis of absolute abundances and for the joint analysis of variation of many species across many samples, as is typical in microbiome association testing, have yet to be investigated.

@@

Here we use a combination of theoretical analysis, simulation, and re-analysis of published experiments to consider when and why taxonomic bias in MGS measurements leads to spurious results in DA analysis of relative and absolute abundances.
We show that, in contrast to received wisdom, taxonomic bias can affect the results of DA methods (for relative and absolute abundance) that are based on species proportions.
We describe the theoretical conditions when this effect is negligible and when it may cause serious scientific errors, and explore this effect in case studies based on real microbiome experiments.
We further demonstrate that another set of DA methods, based on the ratios among species, are robust to bias.
Finally, we present several methods for quantifying, correcting, or otherwise accounting for taxonomic bias in DA analyses which, in many cases, can be deployed with only modest changes to existing experimental and analytical workflows.
Application of these insights and methods can build the confidence needed to turn the findings of microbiome studies into readily-translatable scientific knowledge.

>>>>

<!-- P: Taxonomic bias and DA -->
The standard approach to countering taxonomic bias is to standardize the measurement protocol used within a given study.
Statistical analyses are then conducted with the tacit assumption that all samples will be affected by bias in the same way and so the differences between samples will be unaffected.
This argument is at least intuitively plausible for DA analyses based on multiplicative or fold differences (FDs) in a taxon's abundance.
If bias caused a species' abundance to be consistently measured as 10-fold greater than its actual value, then we would still recover the correct FDs among samples.
<!-- (@kevorkian2018esti, @lloyd2020evid). -->
However, @mclaren2019cons showed mathematically and with MGS measurements of artificially constructed ('mock') communities that consistent taxonomic bias can create fold errors (FEs) that vary across samples and, as a result, majorly distort cross-sample comparisons.
In particular, they showed that the FE in a species' proportion---the most common measure of relative abundance---varies among samples, distorting the observed FDs between samples.
In some cases, bias can even lead to incorrect inferences about the direction of change (for example, by causing a taxon that decreased to appear to increase).
Yet @mclaren2019cons also found that other abundance measures---those based on the ratios among species---have constant FEs and may lead to more robust DA analyses.
The implications of these findings for DA analysis of absolute abundances and for the joint analysis of variation of many species across many samples, as is typical in microbiome association testing, have yet to be investigated.

@@

Here we use a combination of theoretical analysis, simulation, and re-analysis of published experiments to consider when and why taxonomic bias in MGS measurements leads to spurious results in DA analysis of relative and absolute abundances.
We show that, in contrast to received wisdom, taxonomic bias can affect the results of DA methods that are based on species proportions, even if bias is the same across samples.
We describe the theoretical conditions when this effect is negligible and when it may cause serious scientific errors, and explore this effect in case studies based on real microbiome experiments.
We further demonstrate that another set of DA methods, based on the ratios among species, are robust to consistent bias.
Finally, we present several methods for quantifying, correcting, or otherwise accounting for taxonomic bias in DA analyses which, in many cases, can be deployed with only modest changes to existing experimental and analytical workflows.
These insights and methods will aid microbiome data analysts in turning the findings of microbiome studies into readily-translatable scientific knowledge.

### 2022-08-19 solutions.Rmd
Make edits from AW and BC

# Potential solutions {#solutions}

@@

We note two important caveats.
First, in most MGS datasets, any given species is likely to be extremely rare in most samples, such that no sequencing reads are observed.
In order to analyze FDs, DA methods interpret these zero counts as evidence that the true abundance of the species is very small (but still positive).
A common approach to enable the computation of FDs in the presence of zeros is to add a small value, or ‘pseudo-count’, to the read counts, making all values positive.
Unfortunately, this procedure violates bias invariance.
How bias interacts with more sophisticated approaches for handling zeros remains an important open question.
Second, the bias invariance of ratios among species does not extend to additive aggregates of species into higher-order taxa such phyla unless bias is conserved within the species group (@mclaren2019cons).
Multiplicative aggregates of species, recently used for biomarker discovery (@riverapinto2018bala, @quinn2020inte), provide an alternative that remains bias-invariant but may be harder to interpret.

If biological or technical considerations favor a DA method that is not bias invariant, it can be beneficial to also use a ratio-based method as a robustness check.
For example, @hevroni2020seas examined how the within-sample ranks of viral species changed from summer to winter.
Since changes in within-sample ranks can be affected by bias, the authors also considered the (bias-invariant) changes in the centered log ratios (CLR) associated with each species.
They found that the species whose within-sample ranks increased also increased in their CLR value, providing evidence that their initial results were not driven by bias.

@@

_Community calibration controls_ are samples whose species identities and relative abundances are known, either by construction or by characterization with a chosen 'gold standard' or _reference protocol_ (@mclaren2019cons).
Inclusion of these samples along with the primary samples in an MGS experiment can enable researchers to calibrate their MGS measurement by directly measuring and removing the effect of bias.
Calibration can be extended to species not in the controls by imputing the efficiencies of the missing species using phylogenetic relatedness, genetic characteristics (such as 16S copy number), and/or phenotypic properties such as cell-wall structure.

@@

In studies of synthetic communities (like the fungal case study) or of ecosystems dominated by a relatively small number of culturable taxa (like the vaginal microbiome), calibration controls can be artificially constructed ('mock') communities.
In other cases, _natural community controls_ can be derived from aliquots of a homogenized microbiome sample.
Measurement of these controls across multiple experiments enables characterization of the _differential bias_ between experiments (@mclaren2019cons).
Calibration using differential bias can make results directly comparable across studies that used different.

@@

## Choose absolute-abundance methods with more stable FEs

There are many methods for obtaining species (absolute) abundances from the relative abundances measured by MGS, which had previously not been evaluated for their sensitivity to taxonomic bias (@williamson2021amul being a notable exception).
Our theoretical results show that these methods differ in the extent to which taxonomic bias causes the FEs in species abundances to vary across samples.
The effect of bias on absolute DA analysis can be mitigated by choosing methods with more stable FEs, using either of two general approaches.

@@

Consider the debate over whether flow cytometry or 16S qPCR are better methods for measuring total abundance for the purposes of normalizing 16S amplicon sequencing data (@galazzo2020howt, @jian2021comm).
Flow cytometry directly counts cells, whereas 16S qPCR measures the concentration of the 16S gene following DNA extraction.
Thus the 16S qPCR measurement is subject to bias from extraction, copy-number variation, primer binding, and amplification, just like the 16S sequencing measurement.
Although this shared bias likely makes 16S-qPCR less accurate as a measure of total cell concentration, our theory suggests that it can lead to more accurate FDs across samples.

More generally, these observations suggest that for the purposes of performing an absolute DA analysis from amplicon sequencing measurements, the ideal total-abundance measurement may be qPCR the same marker gene, from the same DNA extraction.
Similar reasoning suggests that for shotgun sequencing, the ideal total-abundance measurement is bulk DNA quantification: Shotgun sequencing and bulk DNA quantification are both subject to bias from extraction and variation in genome size.
Optimal use of these pairings requires thoughtful choices during bioinformatics. For example, performing copy-number correction on amplicon read counts prior to multiplication by qPCR measurements would be counter-productive, as it decouples bias in the two measurements. Additionally, the MGS proportions should be computed prior to discarding any unassigned reads, since species that are missing from the given taxonomy database still contribute to the total concentration of marker-gene and/or bulk DNA.
Future experiments should evaluate the extent to which the taxonomic bias of DNA-based total-abundance measurements is stable across samples and shared with the complementary MGS measurement.

@@

A second approach to obtaining species abundances involves normalizing the MGS count for each species to that of a reference species with constant and/or known abundance (Equation \@ref(eq:density-ratio-meas)).
Our theory predicts that taxonomic bias induces constant FEs in the resulting species abundances which do not affect DA results.

@@

Meta-analysis of microbiome samples measured across multiple studies must contend with the fact that different studies typically use different protocols and hence have different taxonomic biases.
These different biases can be explicitly accounted for in a _bias meta-analysis_ which has the potential to improve statistical power as well as interpretability of multi-study DA analyses.
Parametric meta-analysis models include study-specific latent parameters representing 'batch effects'—non-biological differences in the data from each study which can distort the observed biological patterns.
By estimating these 'nuisance parameters' along with the biological parameters of interest (such as the difference in a species' log abundance between conditions), the meta-analysis aims to reduce statistical bias in the biological parameters created by the non-biological differences among studies (@leek2010tack,@wang2019mana).
In a bias-aware meta-analysis, the meta-analysis model is configured so that (some of) the latent parameters correspond to study- and species-specific relative efficiencies.
If taxonomic bias is consistent within but not between studies, then this approach may improve the ability for the meta-analysis to accurately identify DA patterns as compared to meta-analysis methods in which the 'batch effects' do not reflect the multiplicative and compositional nature of taxonomic bias.
The ability to infer and adjust for the differential bias between studies can be increased by measuring one or more shared control samples alongside the experiment-specific samples.

>>>>

# Solutions {#solutions}

@@

We note two important caveats.
First, in most MGS datasets, any given species is likely to be extremely rare in most samples, such that no sequencing reads are observed.
In order to analyze FDs, DA methods interpret these zero counts as evidence that the true abundance of the species is very small (but still positive).
A common approach to enable the computation of FDs in the presence of zeros is to add a small value, or ‘pseudo-count’, to the read counts, making all values positive.
Unfortunately, this procedure violates bias invariance.
How bias interacts with more sophisticated approaches for handling zeros remains an important open question.
Second, the bias invariance of ratios among species does not extend to additive aggregates of species into higher-order taxa such phyla unless bias is conserved within the species group (@mclaren2019cons).
Multiplicative aggregates of species, recently used for biomarker discovery (@riverapinto2018bala, @quinn2020inte), provide an alternative that remains bias-invariant but are harder to interpret.

If biological or technical considerations favor a DA method that is not bias invariant, it can be beneficial to also use a ratio-based method as a robustness check.
For example, @hevroni2020seas examined how the within-sample ranks of viral species changed from summer to winter.
Since changes in within-sample ranks can be affected by bias, the authors also considered the changes in the centered log ratios (CLR) associated with each species.
They found that the species whose within-sample ranks increased also increased in their CLR value, providing evidence that their initial results were not driven by bias.

@@

_Community calibration controls_ are samples whose species identities and relative abundances are known, either by construction or by characterization with a chosen 'gold standard' or _reference protocol_ (@mclaren2019cons, @clausen2022mode).
Inclusion of these samples along with the primary samples in an MGS experiment can enable researchers to calibrate their MGS measurement by directly measuring and removing the effect of bias.
Calibration can be extended to species not in the controls by imputing the efficiencies of the missing species using phylogenetic relatedness, genetic characteristics (such as 16S copy number), and/or phenotypic properties such as cell-wall structure.

@@

In studies of synthetic communities (like the fungal case study) or of ecosystems dominated by a relatively small number of culturable taxa (like the vaginal microbiome), calibration controls can be artificially constructed ('mock') communities.
In other cases, _natural community controls_ can be derived from aliquots of a homogenized microbiome sample.
Measurement of these controls across multiple experiments enables characterization of the _differential bias_ between experiments (@mclaren2019cons).
Calibration using differential bias can make results directly comparable across studies that used different protocols.

@@

## Absolute-abundance methods with more stable FEs

There are many methods for obtaining absolute abundances from the relative abundances measured by MGS, but these methods have not previously been evaluated for their sensitivity to taxonomic bias (@williamson2021amul being a notable exception).
Our theoretical results show that these methods differ in the extent to which taxonomic bias causes the FEs in species abundances to vary across samples.
The effect of bias on absolute DA analysis can be mitigated by choosing methods with more stable FEs, using either of two general approaches.

@@

Consider the debate over whether flow cytometry or 16S qPCR are better methods for measuring total abundance for the purposes of normalizing 16S amplicon sequencing data (@galazzo2020howt, @jian2021comm).
Flow cytometry directly counts cells, whereas 16S qPCR measures the concentration of the 16S gene following DNA extraction.
Thus the 16S qPCR measurement is subject to bias from extraction, copy-number variation, primer binding, and amplification, just like the 16S sequencing measurement.
Although this shared bias likely makes 16S-qPCR less accurate as a measure of total cell concentration, our theory suggests that it will lead to more accurate FDs---and thus more accurate DA results.

More generally, these observations suggest that for the purposes of performing an absolute DA analysis from amplicon sequencing measurements, the ideal total-abundance measurement is qPCR of the same marker gene, from the same DNA extraction.
Similar reasoning suggests that for shotgun sequencing, the ideal total-abundance measurement is bulk DNA quantification: Shotgun sequencing and bulk DNA quantification are both subject to bias from extraction and variation in genome size.
Optimal use of these pairings requires thoughtful choices during bioinformatics. For example, performing copy-number correction on amplicon read counts prior to multiplication by qPCR measurements would be counter-productive, as it decouples bias in the two measurements. Additionally, the MGS proportions should be computed prior to discarding any unassigned reads, since species that are missing from the given taxonomy database still contribute to the total concentration of marker-gene and/or bulk DNA.
Future experiments should evaluate the extent to which the taxonomic bias of DNA-based total-abundance measurements is stable across samples and shared with the complementary MGS measurement.

@@

A second approach to obtaining absolute species abundances involves normalizing the MGS count for each species to that of a reference species with constant and/or known abundance (Equation \@ref(eq:density-ratio-meas)).
Our theory predicts that taxonomic bias induces constant FEs in the abundances obtained by this approach, which will not affect DA results.

@@

Meta-analysis of microbiome samples measured across multiple studies must contend with the fact that different studies typically use different protocols and hence have different taxonomic biases.
These different biases can be explicitly accounted for in a _bias-aware meta-analysis_ which has the potential to improve statistical power as well as interpretability of multi-study DA analyses.
Parametric meta-analysis models include study-specific latent parameters representing 'batch effects'—non-biological differences in the data from each study which can distort the observed biological patterns.
By estimating these 'nuisance parameters' along with the biological parameters of interest (such as the difference in a species' log abundance between conditions), the meta-analysis aims to reduce statistical bias in the biological parameters created by the non-biological differences among studies (@leek2010tack,@wang2019mana).
In a bias-aware meta-analysis, the meta-analysis model is configured so that (some of) the latent parameters correspond to study- and species-specific relative efficiencies.
If taxonomic bias is consistent within but not between studies, then this approach may improve the ability for the meta-analysis to accurately identify DA patterns as compared to meta-analysis methods in which the 'batch effects' do not reflect the multiplicative and compositional nature of taxonomic bias.
The ability to infer and adjust for the differential bias between studies can be increased by measuring one or more shared control samples alongside the experiment-specific samples.
